---
title: 'Final project practical machine learning : Human activity recognition'''
author: "Cheryf LALEYE"
date: "16 mai 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```


###Goal of the project

The goal of your project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did.

###Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

###Data sources

The training data for this project are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv


###Loading packages 

```{r}

library(caret)
library(rattle)
library(corrplot)

```

###Loading and reading data

```{r}

raw.training <- read.table("./pml-training.csv",na.strings=c("NA",""), sep=",", header=T)
raw.testing <- read.table("./pml-testing.csv",na.strings=c("NA",""), sep=",", header=T)

dim(raw.training)
dim(raw.testing)

```

Let's tcheck the structure of our training data : 

```{r}
str(raw.training)

```


The training data set contains 160 variables/predictors and 19622 obsertvations. The test data contains 160 variables/predictors and 20 obsertvations. The outcome to predict here is the variable **classe**.

###Cleaning data 

Firstly, we will clean our data set by removing some meaningless variables. We know that the data which are useful are the data from accelerometers on the belt, forearm, arm and dumbell. So we'll keep only these measurements for our model prediction.

```{r}
  
  VarTokeep <- grep("belt|arm|dumbell|classe",colnames(raw.training))
  cleaned.training.1 <- raw.training[,VarTokeep]
  
  #We'll do the same processing to the testing raw data.
  cleaned.testing.1 <- raw.testing[,VarTokeep]
  
```

Secondly, we'll remove variables/predictors for which we have more than 15% NAs of total observations.

```{r}
  
  ColToRemove <- which(colMeans(is.na(cleaned.training.1)) > 0.15)
  cleaned.training.2 <- cleaned.training.1[,-ColToRemove]
  
  #We'll do the same processing to the testing raw data.
  ColToRemove <- which(colMeans(is.na(cleaned.testing.1)) > 0.15)
  cleaned.testing.2 <- cleaned.testing.1[,-ColToRemove]
  
```

```{r}

dim(cleaned.training.2)
dim(cleaned.testing.2)

rm(cleaned.training.1)
rm(cleaned.testing.1)

```

Set the testing data after cleaning it: 

```{r}
testing <- cleaned.testing.2

```

There's now 40 variables/predictors for both training and test data set.

###Data partition

In the aim to avoid overfitting and a biased out-of-sample error, we'll create a validation set from the cleaned training  data on which, we'll evaluate the model.

```{r}

set.seed(12345)
index <- createDataPartition(y=cleaned.training.2$classe, p=.7, list = F)
training <- cleaned.training.2[index,]
validation <- cleaned.training.2[-index,] 

```

```{r}
dim(training)
dim(validation)
```

```{r}
outcomeIndex <- which(colnames(training)=="classe")

```


###Data analysis 

* Plotting the features distributions

```{r}
featurePlot(training[,-outcomeIndex],training$classe,plot="strip")

```

Since there's no real difference between the distributions of the features, we don't need to perform features scaling (normalization).

* Plotting the correlation between features

```{r}
corrplot(cor(training[,-outcomeIndex]),order="hclust", method="number",addCoef.col="grey")

```

We see tht there are some features which have a strong correlations between each other.

```{r}
StrongCorrelated <- findCorrelation(cor(training[,-outcomeIndex]), cutoff =.7)

```

```{r}
data.frame(names=colnames(training[,StrongCorrelated]),column.position=StrongCorrelated)

```

So maybe, we'll need to perform PCA processing only on these variables in order to :

* reduce noise
* reduce number of predictors
* get uncorrelated (orthogonal) predictors

###Preprocessing

```{r}
#Subset the training set with strong correlated features
train.subset1 <- training[,StrongCorrelated]
validation.subset1 <- validation[,StrongCorrelated]

#Subset the training set without strong correlated features
train.subset2 <- training[,-StrongCorrelated]
validation.subset2 <- validation[,-StrongCorrelated]

```

```{r}
#perform pca on the train.subset and validation.subset
pcaObj.1 <- preProcess(train.subset1, method = "pca", thresh = .95)
train.pca.subset1 <- predict(pcaObj.1, newdata=train.subset1)

pcaObj.2 <- preProcess(validation.subset1, method = "pca", thresh = .95)
validation.pca.subset1 <- predict(pcaObj.2, newdata=validation.subset1)

```

```{r}
dim(train.pca.subset1)
dim(validation.pca.subset1)

```

```{r}
str(train.pca.subset1)

```

Since PCA method perform automatically features scaling by normalizing our variables, we'll need to standardizize our subsets without strong correlated features before pasting it with our pca.subset data.


```{r}
standard.train.obj <- preProcess(train.subset2[,-outcomeIndex], method=c("center","scale"))
standard.validation.obj <- preProcess(validation.subset2[,-outcomeIndex], method=c("center","scale"))

train.subset <- predict(standard.train.obj, newdata=train.subset2[,-outcomeIndex])
validation.subset <- predict(standard.validation.obj, newdata=validation.subset2[,-outcomeIndex])

```

Now we'll paste the features on which we performed PCA and the data subsets just above in order to get our final training.pca and validation.pca datasets.

```{r}
training.pca <- data.frame(train.subset,train.pca.subset1)
validation.pca <- data.frame(validation.subset, validation.pca.subset1)

```

Let's tcheck our training.pca data set structure

```{r}
str(training.pca)

```

###Training data and model selection

Now we'll perform machine learning algorithms on 2 models :

** The model with PCA processing ("training.pca" and "validation.pca")**

* Random forest with the PCA processing performed data

```{r}
#Fitting with random forest
ptm <- proc.time()
model.rf.1 <- train(classe~., data=training.pca, method="rf", trControl = trainControl(method = "cv", 5), ntree = 200)
proc.time() - ptm

```

```{r}
model.rf.1$finalModel

```


```{r}
par(mfrow=c(1,2))
plot(model.rf.1$finalModel,cex=0.7, pch=16, main="Error Vs No. of trees")
varImpPlot(model.rf.1$finalModel, cex=0.7, pch=16, main="Var importance - r.forest (PCA model)")

```

We'll now evaluate the model and get the out-of-sample error on validation set :

```{r}
pred.validation.pca <- predict(model.rf.1, newdata = training.pca)
confusionMatrix(pred.validation.pca, training.pca$classe)
  
```

* Random forest on the data without PCA processing 


```{r}
#Fitting with random forest
ptm <- proc.time()
model.rf <- train(classe~., data=training, method="rf", trControl = trainControl(method = "cv", 5), ntree = 200)
proc.time() - ptm

```

```{r}
pred.validation <- predict(model.rf, newdata = training)
confusionMatrix(pred.validation, training$classe)

```

```{r}
#Fitting with gradient boosting
ptm.gbm <- proc.time()
model.gbm <- train(classe~., data=training.pca, method="gbm", verbose=F)
proc.time() - ptm.gbm

```


```{r}
#Fitting with logistic regression
ptm.glm <- proc.time()
model.glm <- train(classe~., data=training.pca, method="glm")
proc.time() - ptm.glm

```






* The model without PCA processing ("training" and "validation")


